Exploring Axiom Systems for Category Theory

### 2017-02-10 FreeLogic/2017-Journal/article.tex
added

\author{Christoph Benzmüller       \and
      Dana Scott 
}

@@

In this article we show how free logic can be
``implemented'' in any theorem proving system for classical
higher-order logic (HOL) \cite{B5}. The proposed solution employs a
semantic embedding of free (or inclusive logic) in HOL. We present an
exemplary implementation of this idea in the mathematical proof
assistant Isabelle/HOL \cite{NPW02}. Various state-of-the-art
first-order and higher-order automated theorem provers and model
finders are integrated (modulo suitable logic translations) with
Isabelle via the Sledgehammer tool \cite{Sledgehammer}, so that our
solution can be utilized, via Isabelle as foreground system, with a
whole range of other background reasoners. As a result we obtain an
elegant and powerful implementation of an interactive and automated
theorem proving (and model finding) system for free logic.

  The two main contributions of this article are: (i) A shallow
  semantical embedding of free logic in classical higher-order logic
  is presented; this embedding enables the application of higher-order
  interactive and automated theorem provers (and their integrated
  subprovers) for the formalisation and verification of free logic
  theories.  (ii) This approach to automate free logic is exemplarily
  employed in a selected domain of mathematics: Starting from a
  generalization of the standard axioms for a monoid we present a
  stepwise development of various, mutually equivalent foundational
  axiom systems for category theory. As a side-effect of this work
  some (minor) issue in a prominent category theory textbook have been
  revealed.

Partiality and undefinedness are core concepts in various areas of
mathematics.  Modern mathematical proof assistants and theorem proving
systems are often based on traditional classical or intuitionistic
logics and provide rather inadequate support for these challenge
concepts.  Free logic @{cite "sep-logic-free,scott67:exist"}, in
contrast, offers a theoretically and practically appealing
solution. Unfortunately, however, we are not aware of any implemented
and available theorem proving system for free logic.

In this extended abstract we show how free logic can be
``implemented'' in any theorem proving system for classical
higher-order logic (HOL) @{cite "B5"}. The proposed solution employs a
semantic embedding of free (or inclusive logic) in HOL. We present an
exemplary implementation of this idea in the mathematical proof
assistant Isabelle/HOL @{cite "NPW02"}. Various state-of-the-art
first-order and higher-order automated theorem provers and model
finders are integrated (modulo suitable logic translations) with
Isabelle via the Sledgehammer tool @{cite "Sledgehammer"}, so that our
solution can be utilized, via Isabelle as foreground system, with a
whole range of other background reasoners. As a result we obtain an
elegant and powerful implementation of an interactive and automated
theorem proving (and model finding) system for free logic.

To demonstrate the practical relevance of our new system, we present a
stepwise development of axiom systems for category theory by
generalizing the standard axioms for a monoid to a partial composition
operation. Our purpose is not to make or claim any contribution to
category theory but rather to show how formalizations involving the
kind of logic required (free logic) can be validated within modern
proof assistants.

A total of eight different axiom systems is studied. The systems I-VI are shown to 
be equivalent. The axiom system VII slightly modifies axiom system VI to obtain (modulo 
notational transformation) the set of axioms as proposed by  Freyd and Scedrov in their textbook
 ``Categories, Allegories'' @{cite "FreydScedrov90"}, published in 1990; 
see also Subsection \ref{subsec:FreydNotation} where we present their original system.
While the axiom systems I-VI are shown to be  consistent, a constricted inconsistency result is 
obtained for system VII (when encoded in free logic where free variables range over all 
objects): We can prove @{text "(∃x. ❙¬(E x)) ❙→ False"}, where @{text "E"} is the existence predicate. Read this as: If there 
are undefined objects, e.g. the value of an undefined composition @{text "x⋅y"}, then we have falsity.
By contraposition, all objects (and thus all compositions) must exist. But when we assume the latter,
then the axiom system VII essentially reduces categories to monoids.
We note that axiom system V, which avoids this problem, corresponds to a set of axioms proposed 
by Scott @{cite "Scott79"} in the 1970s. The problem can also be avoided by restricting the variables 
in axiom system VII to range only over existing objects and by postulating strictness conditions. 
This gives us axiom system VIII.

@@

To enable our experiments we have exploited an embedding of free logic @{cite "Scott67"} 
in classical higher-order logic, which we have recently presented in a related paper @{cite "C57"}.

We also want to emphasize that this paper has been written entirely within the Isabelle 
framework by utilizing the Isabelle ``build'' tool; cf. @{cite "IsabelleManual2016"}, Section~2. 
It is thus an example of a formally verified mathematical document, where the PDF document as 
presented here has been generated directly from the verified source files mentioned above.
We also note that once the proofs have been mechanically checked, they are generally easy 
to find by hand using paper and pencil.

@@

% Non-BibTeX users please use
\begin{thebibliography}{}
%
% and use \bibitem to create references. Consult the Instructions
% for authors for reference list style.
%
\bibitem{RefJ}
% Format for Journal Reference
Author, Article title, Journal, Volume, page numbers (year)
% Format for books
\bibitem{RefB}
Author, Book title, page numbers. Publisher, place (year)
% etc
\end{thebibliography}

>>>>

\author{Christoph Benzm\"uller       \and
      Dana Scott 
}

@@

In the first part of this article (\S2 and \S3) we show how free logic can be
``implemented'' in any theorem proving system for classical
higher-order logic (HOL) \cite{B5}. The proposed solution employs a
semantic embedding of free (or inclusive logic) in HOL. We present an
exemplary implementation of this idea in the mathematical proof
assistant Isabelle/HOL \cite{NPW02}. Various state-of-the-art
first-order and higher-order automated theorem provers and model
finders are integrated (modulo suitable logic translations) with
Isabelle via the Sledgehammer tool \cite{Sledgehammer}, so that our
solution can be utilized, via Isabelle as foreground system, with a
whole range of other background reasoners. As a result we obtain an
elegant and powerful implementation of an interactive and automated
theorem proving (and model finding) system for free logic.

% Partiality and undefinedness are core concepts in various areas of
% mathematics.  Modern mathematical proof assistants and theorem proving
% systems are often based on traditional classical or intuitionistic
% logics and provide rather inadequate support for these challenge
% concepts.  Free logic @{cite "sep-logic-free,scott67:exist"}, in
% contrast, offers a theoretically and practically appealing
% solution. Unfortunately, however, we are not aware of any implemented
% and available theorem proving system for free logic.

% In this extended abstract we show how free logic can be
% ``implemented'' in any theorem proving system for classical
% higher-order logic (HOL) @{cite "B5"}. The proposed solution employs a
% semantic embedding of free (or inclusive logic) in HOL. We present an
% exemplary implementation of this idea in the mathematical proof
% assistant Isabelle/HOL @{cite "NPW02"}. Various state-of-the-art
% first-order and higher-order automated theorem provers and model
% finders are integrated (modulo suitable logic translations) with
% Isabelle via the Sledgehammer tool @{cite "Sledgehammer"}, so that our
% solution can be utilized, via Isabelle as foreground system, with a
% whole range of other background reasoners. As a result we obtain an
% elegant and powerful implementation of an interactive and automated
% theorem proving (and model finding) system for free logic.

To demonstrate the practical relevance of our new system, we present,
in the second part of this article, a stepwise development of axiom
systems for category theory by generalizing the standard axioms for a
monoid to a partial composition operation. Our purpose is not to make
or claim any contribution to category theory but rather to show how
formalizations involving the kind of logic required (free logic) can
be validated within modern proof assistants.

A total of eight different axiom systems is studied. The systems I-VI
are shown to be equivalent. The axiom system VII slightly modifies
axiom system VI to obtain (modulo notational transformation) the set
of axioms as proposed by Freyd and Scedrov in their textbook
``Categories, Allegories'' \cite{FreydScedrov90}, published in 1990;
see also Subsection \ref{subsec:FreydNotation} where we present their
original system.  While the axiom systems I-VI are shown to be
consistent, a constricted inconsistency result is obtained for system
VII (when encoded in free logic where free variables range over all
objects): We can prove $\exists x. \neg(E x) \rightarrow False$, where
$E$ is the existence predicate. Read this as: If there are undefined
objects, e.g. the value of an undefined composition $x\cdot y$
then we have falsity.  By contraposition, all objects (and thus all
compositions) must exist. But when we assume the latter, then the
axiom system VII essentially reduces categories to monoids.  We note
that axiom system V, which avoids this problem, corresponds to a set
of axioms proposed by Scott @{cite "Scott79"} in the 1970s. The
problem can also be avoided by restricting the variables in axiom
system VII to range only over existing objects and by postulating
strictness conditions.  This gives us axiom system VIII.

@@

The content of article is combines, extends and clarifies the
contributions reported in two previous papers \cite{ICMS,ArXiv}.

% To enable our experiments we have exploited an embedding of free logic @{cite "Scott67"} 
% in classical higher-order logic, which we have recently presented in a related paper @{cite "C57"}.

% We also want to emphasize that this paper has been written entirely within the Isabelle 
% framework by utilizing the Isabelle ``build'' tool; cf. @{cite "IsabelleManual2016"}, Section~2. 
% It is thus an example of a formally verified mathematical document, where the PDF document as 
% presented here has been generated directly from the verified source files mentioned above.
% We also note that once the proofs have been mechanically checked, they are generally easy 
% to find by hand using paper and pencil.

\section{Preliminaries}
\label{sec:preliminaries}
\subsection{Free Logic}
Free logic \cite{Lambert60,Scottt} refers to a class of logic
formalisms that are free of basic existence assumptions regarding the
denotation of terms. Remember that terms in e.g. traditional classical
and intuitionistic predicate logics always denote an object in a given
(non-empty) domain $D$, which also serves as the domain of
quantification, that is, the set of objects the existential and
universal quantifiers of the logic range over. In free logic these
basic assumption are abolished. Terms do still denote objects in a
(non-empty) domain $D$, but a (possibly empty) set $E\subseteq D$ is
added to characterize the subdomain of ``existing'' resp. ``defined''
objects in $D$. Quantification, in contrast to traditional logics, now
ranges over this set $E$ of existing/defined objects only. It is
obvious how this helps to model undefideness and partiality:
problematic terms, such as $1/0$, still denote, but they refer to
undefined objects, that is, objects $d$ in $D\setminus E$ lying
outside of the scope of quantification.

@@

% % Non-BibTeX users please use
% \begin{thebibliography}{}
% %
% % and use \bibitem to create references. Consult the Instructions
% % for authors for reference list style.
% %
% \bibitem{RefJ}
% % Format for Journal Reference
% Author, Article title, Journal, Volume, page numbers (year)
% % Format for books
% \bibitem{RefB}
% Author, Book title, page numbers. Publisher, place (year)
% % etc
% \end{thebibliography}
